Current and future constraints on primordial magnetic fields

### 2017-02-03 pmf.tex
to 4.3.2

\newcommand{\sqdeg}{deg$^2$ }
\newcommand{\omb}{\ensuremath{\Omega_b h^2}}
\newcommand{\omc}{\ensuremath{\Omega_c h^2}}
\newcommand{\clpp}{\ensuremath{C_{L}^{\phi\phi}}}
\newcommand{\cpmf}{\ensuremath{C_{\ell}^{\rm PMF}}}
\newcommand{\apmf}{\ensuremath{A_{\rm PMF}}}
\newcommand{\bpmf}{\ensuremath{B_{\rm 1\,Mpc}}}
\newcommand{\alens}{\ensuremath{A_{\rm lens}}}
\newcommand{\lcdm}{\ensuremath{\Lambda}CDM}
\newcommand{\nrun}{\ensuremath{n_{\rm run}}}
\newcommand{\neff}{\ensuremath{N_{\rm eff}}}
\newcommand{\ho}{H\ensuremath{_0}}
\newcommand{\mnu}{\ensuremath{\sum m_\nu}}
\newcommand{\ukarcmin}{\ensuremath{\mu}K-arcmin}
\newcommand{\lknee}{\ensuremath{\ell_{\rm knee}}}
\newcommand{\fermilat}{\textit{Fermi}-LAT}

@@

The experiments that will begin taking data in 2017 will dramatically improve constraints on the PMF power. 
With the minimal cosmological model, the 1-$\sigma$ forecasts for \planck{}+\ho{} is $\sigma(\apmf)=\fisherAplk$ for a nearly-scale invariant PMF spectrum (i.e.~$n_B=-2.9$). 
This is consistent with, although somewhat better than, the upper limits from the real data).  
Adding the EE and BB bandpowers from the two stage III experiments (\sptnew{} and \simons{}) reduces the uncertainty by an order of magnitude in the 7-parameter model to $\sigma(\apmf)=0.015$. 
The improvement is larger when including all eleven parameters, as the parameter degeneracies substantially weaken (by a factor of five) the \planck+\ho{} constraints on \apmf{}, while only weakening the stage III constraint by a  more modest 40\%. 
Thus within the 11-parameter model however,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,35 to  $\sigma(\apmf)=0.022$. 
The parameter degeneracies largely disappear for steeper PMF spectra (i.e. $n_B=-1$ or 2) as the PMF B-mode spectra then peaks at very small scales and this small-scale power can not be mimicked by any of the other 11-parameters. 
The improvement from adding the SPT-3G and Simons Array experiments to \planck{} remains impressive.
For the 11-parameter model and $n_B=-1$, adding SPT-3G and Simons Array to \planck{} leads to a 17-fold reduction in $\sigma(\apmf)$ from \fisherAlooseNbOneplk{} to \fisherAlooseNbOneSThree{}.
For $n_B=2$, there is a 20-fold reduction from $\sigma(\apmf) = \fisherAlooseNbTwoplk$ to  \fisherAlooseNbTwoSThree. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivation behind the proposed CMB stage-IV experiment commonly called CMB-S4 is to search for inflationary gravitational waves and neutrino masses.
However, CMB-S4 would also enable extremely sensitive searches for PMFs. 
We begin by considering our fiducial CMB-S4 configuration, as laid out in Table~\ref{tab:experiments}. 
For this fiducial configuration and the 11-parameter model, we find a 3-fold reduction over the stage III experiments for all three PMF template shapes. 
The exact values are $\sigma(\apmf^{n_B=-2.9}) =   \fisherAlooseSFour$, $\sigma(\apmf^{n_B=-1}) =   \fisherAlooseNbOneSFour$, and $\sigma(\apmf^{n_B=2}) =   \fisherAlooseNbTwoSFour$. 
These represent more than a 50-fold improvement on current limits.

Given that CMB-S4 is currently in the design phase, it is also worthwhile to consider how design decisions will affect the final PMF constraints. 
We look at five aspects of the experiment design: the instrumental sensitivity (as reflected by the map noise level), the telescope size or beam FWHM,  the low-$\ell$ noise performance (1/f knee), the choice of survey area,  and how well the beam size and calibration must be known. 
Of these five, we find that the first two to be very important and the second two to be somewhat important when optimizing for searches for PMFs. 
As should be expected, the preferred design depends on the shape of the PMF spectrum ($n_B$) which controls  the relative power between large and small angular scales. 
However, as was discussed in \S\ref{sec:template},  there are theoretical reasons to expect the PMF spectrum to be close to a scale-invariant spectra, as opposed to, e.g., $n_B=2$.

@@

Reducing the noise in CMB maps should always improve the derived constraints. 
The goal of this section is to quantify the magnitude of that improvement: has the information gain saturated, or will further reductions in noise substantially improve PMF searches. 
Figure~\ref{fig:sensitivity} shows the uncertainty on \apmf{} as a function of the map noise level for the three PMF templates in either the 11 or 7-parameter cosmological models. 
For this figure, we have fixed the rest of the experimental configuration (i.e.~everything except the polarized noise levels) to the values of the CMB-S4 row in Table~\ref{tab:experiments}. 
Clearly, continuing to improve the sensitivity of CMB experiments to CMB-S4 and beyond will be a boon to PMF searches. 
In the general, 11-parameter cosmological model, the PMF uncertainty does not plateau in any of the three PMF templates considered till the map noise is at or below $0.3\,\ukarcmin$ (a factor of four lower then the fiducial CMB-S4 configuration).

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_sens.pdf}
  \caption[]{ \label{fig:sensitivity}
  How the expected uncertainty on the PMF power scales with the polarized map noise. 
  Forecasts are shown for each of the three PMF scalar indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
  In the case of $n_B=2$, the uncertainties have been multiplied by a factor of $10^5$ to allow them to be plotted on the same scale. 
  The black lines are for the conservative 11-parameter model, while the light blue lines are for the simple \lcdm{}+\apmf{} model. 
  The model only matters in the nearly-scale invariant PMF case, where the constraints degrade in the 11-parameter model due a degeneracy with the tensor-to-scalar ratio. 
  The PMF constraints improve rapidly with lower noise levels in all three cases up to $\sim$\,1\,\ukarcmin, which is close to the  fiducial CMB-S4 noise levels. 
    }
\end{figure}

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_beam.pdf}
  \caption[]{ \label{fig:beam}
 Larger telescopes (and smaller beams) will improve limits on the PMF power. 
 Forecasts are shown for each of the three PMF scalar indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
   In the $n_B=2$ case, the uncertainties have been multiplied by a factor of $10^5$ to allow them to be plotted on the same scale. 
  The black lines are for the conservative 11-parameter model, while the light blue lines are for the simple \lcdm{}+\apmf{} model. 
  The cosmological model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size is relatively unimportant. 
  The plateau near 4$^\prime$ for $n_B=-1, +2$ is where the gain from the $\ell\sim3000$ peak is saturated, but the resolution isn't yet adequate to pick up the next peak at $\ell > 8000$ (see  Fig.~\ref{fig:pmf-nb}).  
    }
\end{figure}

Larger telescopes will allow CMB polarization to be measured on smaller angular scales, and should therefore always improve the PMF constraints. 
As shown in Fig.~\ref{fig:beam}, we find the gains due to resolution at fixed mapping speed to be substantial for all three power law indices considered. 
We find a local plateau between FWHMs of 4$^\prime$ to 6$^\prime$; improving the angular resolution in this range does not aid PMF searches. 
We can understand this plateau by looking on the predicted PMF power in Fig.~\ref{fig:pmf-nb}. 
All three templates have two peaks across this range of angular multiples, and we would expect such a plateau when, for instance, the resolution is adequate to resolve the first peak, but not yet sufficient to resolve the second peak. 
The observed plateau positions are consistent with this hypothesis. 
 A beam size of FWHM=1$^\prime$ instead of 10$^\prime$ reduces the expected upper limit five-fold for $n_B=-2.9$ and 30- to 35-fold for the steeper PMF indices which are more heavily weighted towards small angular scales. 
   The model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size is relatively unimportant.

However, larger telescopes are also more expensive to build which means for a fixed experimental budget, they would necessitate less ambitious focal planes and lower instantaneous mapping speed. 
We very crudely approximated a cost-neutral setup with a noise level of 1.3 \ukarcmin{} for a FWHM of 4$^\prime$; a noise level of 2.9 \ukarcmin{} for a FWHM of 2$^\prime$; and a noise level of 4.1 \ukarcmin{} for a FWHM of 1$^\prime$. 
We find 
Improving the mapping speed rather than telescope size leads to somewhat better potential limits for these experimental configurations, by a factor of 1.6, 1.2 or 1.1 for $n_B = -2.9, -1, +2$ respectively. 
However the results are close enough that the crudeness of the cost-neutral estimates used here could be a factor, and the preference might flip for more accurately costed setups. 
In short,  the tradeoff between larger telescopes or more detectors is largely a wash, although one would likely lean towards adding more detectors. 
%Raw sensitivity is important to an experiment's ability to study PMFs.

>>>>

\newcommand{\sqdeg}{deg$^2$ }
\newcommand{\omb}{\ensuremath{\Omega_b h^2}}
\newcommand{\omc}{\ensuremath{\Omega_c h^2}}
\newcommand{\clpp}{\ensuremath{C_{L}^{\phi\phi}}}
\newcommand{\cpmf}{\ensuremath{C_{\ell}^{\rm PMF}}}
\newcommand{\apmf}{\ensuremath{A_{\rm PMF}}}
\newcommand{\bpmf}{\ensuremath{B_{\rm 1\,Mpc}}}
\newcommand{\alens}{\ensuremath{A_{\rm lens}}}
\newcommand{\lcdm}{\ensuremath{\Lambda}CDM}
\newcommand{\nrun}{\ensuremath{n_{\rm run}}}
\newcommand{\neff}{\ensuremath{N_{\rm eff}}}
\newcommand{\ho}{H\ensuremath{_0}}
\newcommand{\mnu}{\ensuremath{\sum m_\nu}}
\newcommand{\ukarcmin}{\ensuremath{\mu}{\rm K-arcmin}}
\newcommand{\lknee}{\ensuremath{\ell_{\rm knee}}}
\newcommand{\fermilat}{\textit{Fermi}-LAT}

@@

The experiments that will begin taking data in 2017 will dramatically improve our knowledge of PMFs. 
In the minimal cosmological model, the 1-$\sigma$ forecast for \planck{}+\ho{} is $\sigma(\apmf)=\fisherAplk$ for a best-motivated, nearly-scale invariant PMF spectrum (i.e.~$n_B=-2.9$). 
This is on-par with, although 30\% better than, the actual upper limit from the \planck{} data.  
Adding  EE and BB bandpowers from two stage-III experiments, \sptnew{} and \simons{}, reduces the uncertainty by an order of magnitude in the 7-parameter model to $\sigma(\apmf)=0.015$. 
The relative improvement is larger in the more realistic 11-parameter model, as parameter degeneracies substantially weaken (by a factor of five) the \planck+\ho{} constraints on \apmf{}, while weakening the stage III constraint by a  more modest 40\%. 
Thus within the 11-parameter model however,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,35 to  $\sigma(\apmf)=0.022$. 
The parameter degeneracies largely disappear for steeper PMF spectra (i.e. $n_B=-1$ or 2) as the PMF B-mode spectra then peaks at very small scales and this small-scale power can not be mimicked by any of the other 11-parameters. 
The improvement from adding the SPT-3G and Simons Array experiments to \planck{} remains impressive for these values of $n_B$.
For the 11-parameter model and $n_B=-1$, adding SPT-3G and Simons Array to \planck{} leads to a 17-fold reduction in $\sigma(\apmf)$ from \fisherAlooseNbOneplk{} to \fisherAlooseNbOneSThree{}.
For $n_B=2$, there is a 20-fold reduction from $\sigma(\apmf) = \fisherAlooseNbTwoplk$ to  \fisherAlooseNbTwoSThree. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivations behind the proposed CMB-S4 experiments are to search for inflationary gravitational waves and to measure the neutrino masses.
However, CMB-S4 would also enable extremely sensitive searches for PMFs. 
We begin by considering our fiducial CMB-S4 configuration, as laid out in Table~\ref{tab:experiments}. 
For this fiducial configuration and the 11-parameter model, we find a 3-fold reduction over the stage III experiments for all three PMF template shapes. 
The expected uncertainties for CMB-S4 are $\sigma(\apmf^{n_B=-2.9}) =   \fisherAlooseSFour$, $\sigma(\apmf^{n_B=-1}) =   \fisherAlooseNbOneSFour$, and $\sigma(\apmf^{n_B=2}) =   \fisherAlooseNbTwoSFour$. 
These represent more than a 50-fold improvement on current limits.

Given that CMB-S4 is being designed, it is worth considering how design decisions would affect the final PMF constraints. 
We look at five aspects of the experiment: the instrumental sensitivity (as reflected by the final noise level in the maps), the telescope size or beam FWHM,  the low-$\ell$ noise performance, the amount of sky surveyed,  and how well the beam shape and calibration must be known. 
Of these five aspects, we find the first two to be very important to PMF searches and the second two to be somewhat important. 
As might be expected, in some cases the optimal design depends on the shape of the PMF spectrum ($n_B$) as this sets  the relative power between large and small angular scales.  
However, as was discussed in \S\ref{sec:template},  there are theoretical reasons to expect the PMF spectrum to be close to a scale-invariant spectra. 
While for completeness we present results for three values of $n_B$, we recommend favoring the optimum for $n_B=-2.9$.

@@

Reducing the noise in CMB maps will monotonically improve power spectrum measurements. 
The goal of this section is to quantify the magnitude of that improvement: has the information gain saturated, or will further reductions in noise substantially improve PMF searches. 
Figure~\ref{fig:sensitivity} shows the uncertainty on \apmf{} as a function of the map noise level for the three PMF templates in either the 11 or 7-parameter cosmological models. 
For this figure, we have fixed the rest of the experimental configuration (i.e.~everything except the polarized noise levels) to the values of the CMB-S4 row in Table~\ref{tab:experiments}. 
Clearly, continuing to improve the sensitivity of CMB experiments to CMB-S4 and beyond will be a major boon to PMF searches. 
In the general, 11-parameter cosmological model, the PMF uncertainty does not plateau in any of the three PMF templates considered till the map noise is at or below $0.3\,\ukarcmin$ (a factor of four lower then the fiducial CMB-S4 configuration).

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_sens.pdf}
  \caption[]{ \label{fig:sensitivity}
  How the PMF uncertainty  scales with the polarized map noise. 
  Forecasts are shown as a function of the map noise in \ukarcmin{} for each of the three PMF spectral indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
  In the case of $n_B=2$, the uncertainties have been multiplied by a factor of $10^5$ to allow them to be plotted on the same scale. 
  The black lines are for the fiducial 11-parameter model, while the light blue lines are for the minimal \lcdm{}+\apmf{} model. 
  The model only matters in the nearly-scale invariant PMF case, where the constraints degrade in the 11-parameter model due a degeneracy with the tensor-to-scalar ratio. 
  The PMF constraints improve rapidly with lower noise levels in all six cases up to $\sim$\,1\,\ukarcmin, which is close to the  fiducial CMB-S4 noise levels. 
    }
\end{figure}

@@

\begin{figure}[htb]\centering
\includegraphics[width=0.45\textwidth,clip,trim={2.cm 12.5cm 11cm 7.5cm}]{pmf_beam.pdf}
  \caption[]{ \label{fig:beam}
 Larger telescopes (and thus smaller beams) improve limits on the PMF power. 
 Forecasts are shown as a function of the beam FWHM in arcminutes for each of the three PMF scalar indices: $n_B = -2.9$ (solid), -1 (dashed), and +2 (dash-dot). 
   In the $n_B=2$ case, the uncertainties have been multiplied by a factor of $10^5$ for plotting purposes. 
  The black lines are for the fiducial 11-parameter model, while the light blue lines are for the minimal \lcdm{}+\apmf{} model. 
  The cosmological model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size is relatively unimportant. 
  The plateau near 4$^\prime$ for $n_B=-1, +2$ is where the gain from the $\ell\sim3000$ peak is saturated, but the resolution isn't yet adequate to pick up the next peak at $\ell > 8000$ (see  Fig.~\ref{fig:pmf-nb}).  
    }
\end{figure}

Larger telescopes can recover more modes on the sky, and should therefore always improve the PMF constraints. 
As shown in Fig.~\ref{fig:beam}, we find the gains due to resolution at fixed mapping speed to be substantial for all three power law indices considered. 
We see a local plateau between FWHMs of 4$^\prime$ to 6$^\prime$; improving the angular resolution in this range does little to aid PMF searches. 
We can understand this plateau by looking on the predicted PMF power in Fig.~\ref{fig:pmf-nb}. 
All three templates have two peaks across this range of angular multiples, and we would expect to see such a plateau when, for instance, the resolution is adequate to resolve the first peak, but not yet sufficient to resolve the second peak. 
The observed plateau positions are consistent with this hypothesis. 
 Overall, a beam size of FWHM=1$^\prime$ instead of 10$^\prime$ reduces the expected upper limit five-fold for $n_B=-2.9$ and 30- to 35-fold for the steeper PMF indices which are more heavily weighted towards small angular scales. 
   The model only matters in the nearly-scale invariant PMF case ($n_B=-2.9$), where the information on large angular scales dominates and the telescope size is relatively unimportant.

However, larger telescopes are also more expensive to build which means for a fixed experimental budget, they would necessitate less ambitious focal planes and lower instantaneous mapping speeds. 
We very crudely approximated a cost-neutral setup with a noise level of 1.3 \ukarcmin{} for a FWHM of 4$^\prime$; a noise level of 2.9 \ukarcmin{} for a FWHM of 2$^\prime$; and a noise level of 4.1 \ukarcmin{} for a FWHM of 1$^\prime$. 
We find that 
improving the mapping speed rather than telescope size leads to somewhat better potential limits for these experimental configurations, by a factor of 1.6, 1.2 or 1.1 for $n_B = -2.9, -1, +2$ respectively. 
However the results are close enough that the crudeness of the cost-neutral estimates used here could be a factor, and the preference might flip for more accurately costed setups. 
In short,  the tradeoff between larger telescopes or more detectors is largely a wash, although one might lean towards adding more detectors. 
%Raw sensitivity is important to an experiment's ability to study PMFs.
